

\bibliographystyle{plainnat}

\begin{thebibliography}{10}

\bibitem{Higham2008}
N. J. Higham.
\newblock Functions of Matrices.
\newblock {\em Siam}, ISBN: 978-0-89871-646-7, 2008.

\bibitem{BT1998}
L. Brugnano and D. Trigiante.
\newblock Solving Differential Problems by Multistep Initial and Boundary Value Methods.
\newblock {\em Australia Etc. : Gordon \& Breach}, 1998.

\bibitem{LT2002}
V. Lakshmikantham and D. Trigiante.
\newblock Theory of Difference Equations : Numerical Methods and Applications.
\newblock {\em  New York : Marcel Dekker}, 2002.

\bibitem{Gantmacher1959}
F. R. Gantmacher.
\newblock The Theory of Matrices, volume one.
\newblock New York, 1959.

\bibitem{GL1996}
G. H. Golub and C. F. Van Loan.
\newblock Matrix Computations.
\newblock {\em Johns Hopkins University Press},
    Baltimore, MD, USA, third edition, 1996.

\bibitem{HJ1991}
R. A. Horn and C. R. Johnson.
\newblock Topics in Matrix Analysis.
\newblock {\em Cambridge University Press}, 1991.

\bibitem{LT1985}
P. Lancaster and M. Tismenetsky.
\newblock The Theory of Matrices, second edition.
\newblock {\em Academic Press, London}, 1985.


\bibitem{OEIS}
OEIS~Foundation Inc.
\newblock The On-Line {E}ncyclopedia of {I}nteger {S}equences.
\newblock \url{https://oeis.org/}.

\bibitem{MRSV97}
D.Merlini, D.~G. Rogers, R.~Sprugnoli and M.~C. Verri.
\newblock On some alternative characterizations of {R}iordan arrays.
\newblock {\em Canadian Journal of Mathematics}, 49(2):301--320, 1997.

\bibitem{SGWW91}
L.~W. Shapiro, S.~Getu, W.-J. Woan and L.~Woodson.
\newblock The {R}iordan group.
\newblock {\em Discrete Applied Mathematics}, 34:229--239, 1991.

\bibitem{Spr94}
R. Sprugnoli.
\newblock Riordan arrays and combinatorial sums.
\newblock {\em Discrete Mathematics}, 132:267--290, 1994.

\bibitem{RUNCKEL1983161}
H. Runckel and U. Pittelkow.
\newblock Practical computation of matrix functions.
\newblock {\em Linear Algebra and its Applications}, volume 49, 161-178, 1983.

\bibitem{VERDESTAR2005285}
L. Verde-Star.
\newblock Functions of matrices.
\newblock {\em Linear Algebra and its Applications}, volume 406, 285-300, 2005.

\bibitem{HE2016208}
T.-X. He.
\newblock Applications of Riordan matrix functions to Bernoulli and Euler polynomials.
\newblock {\em Linear Algebra and its Applications}, volume 507, 208-228, 2016.

\bibitem{HE201515}
T.-X. He.
\newblock Matrix characterizations of Riordan arrays.
\newblock {\em Linear Algebra and its Applications}, volume 465, 15-42, 2015.

\bibitem{MSV07}
D. Merlini and R. Sprugnoli and M. C. Verri.
\newblock The method of coefficients.
\newblock {\em American Mathematical Monthly}, 114:40--57, 2007.

\bibitem{LUZON2016239}
A. Luz\'on, D. Merlini, M. A. Mor\'on, L. F. Prieto-Martinez and R. Sprugnoli.
\newblock Some inverse limit approaches to the Riordan group.
\newblock {\em Linear Algebra and its Applications}, 491:239--262, 2016.

\bibitem{CHEON200149}
G.-S. Cheon and J.-S. Kim.
\newblock Stirling matrix via Pascal matrix.
\newblock {\em Linear Algebra and its Applications}, 329:49--59, 2001.

\bibitem{MOLERLOAN2003}
C. Moler and C. Van Loan.
\newblock Nineteen Dubious Ways to Compute the Exponential of a Matrix, Twenty-Five Years Later.
\newblock {\em SIAM Review}, 45:3--49, 2003.



\end{thebibliography}

