
\section{Introduction}

\label{sec:introduction}
\input{Riordan-matrices-functions/introduction.tex}

\section{Basic definitions and notations}

\input{Riordan-matrices-functions/matrix-functions.tex}

We are now ready to apply this framework to matrices in the Riordan group.

\section{Riordan matrices}

\input{Riordan-matrices-functions/riordan-group.tex}

\iffalse % \subsection{A Riordan array characterization of Hermite interpolating polynomials} {{{

\input{Riordan-matrices-functions/Ra-g-characterization.tex}

\subsection{A component matrices characterization of Hermite interpolating polynomials}

\input{Riordan-matrices-functions/Zs-g-characterization.tex}
\fi
% }}}

\section{Functions and polynomials}

In this section we instantiate the abstract framework just described to functions
\begin{displaymath}
\begin{split}
f(z)&=z^{r},\,{f(z)=\frac{1}{z}},\,{f(z)=\sqrt{z}},\,{f(z)=e^{\alpha z}},\\
f(z)&=log\,{z},\,f(z)=sin\,{z}\quad\text{and}\quad f(z)=cos\,{z},
\end{split}
\end{displaymath}
where $r,\alpha\in\mathbb{R}$; in parallel, we construct and show corresponding
Hermite interpolating polynomials in a sequence of theorems, respectively.
From now on, we use $m$ and $\lambda$ instead of $m_{1}$ and $\lambda_{1}$ to
simplify the notation; moreover, we instantiate $\lambda=1$ which is the
natural eigenvalue for Riordan arrays.

We start by generalizing the $r$-th power $A^{r}$, usually carried out
as $\underbrace{A\cdots A}_{r\text{ times}}$, to \textit{non-naturals} powers
$r\in\mathbb{Q}$.

\input{Riordan-matrices-functions/function-pow.tex}

Instantiation $r=-1$ in the previous theorem yields a Hermite interpolating
polynomial for the inverse function which, in the explicit form, reduces to
a binomial transform.

\input{Riordan-matrices-functions/function-inverse.tex}

Instantiation $r=\frac{1}{2}$ yields the interpolation of the square root function,
we report its derivation for completeness.

\input{Riordan-matrices-functions/function-sqrt.tex}

Matrix exponentiation is a well studied problem \citep{MOLERLOAN2003}, here
we show another way in the Riordan arrays domain.

\input{Riordan-matrices-functions/function-exp.tex}

We show a dual theorem of the previous one concerning the interpolation of the
logarithm function.

\input{Riordan-matrices-functions/function-log.tex}

Finally, we show two theorem concerning trigonometric functions $\sin$ and
$\cos$, respectively.

\input{Riordan-matrices-functions/function-sin.tex}

\input{Riordan-matrices-functions/function-cos.tex}

\vfill

\subsection{Case studies}

In this section we apply the functions described in the previous arguments to
Riordan arrays $\mathcal{P}_{8}, \mathcal{C}_{8}$ and $\mathcal{S}_{8}$
concerning binomial coefficients, Catalan and Stirling numbers, respectively.

\input{Riordan-matrices-functions/matricial-Hermite-polys.tex}

\input{Riordan-matrices-functions/case-study-pascal.tex}

\input{Riordan-matrices-functions/case-study-catalan.tex}

\input{Riordan-matrices-functions/case-study-stirling-II.tex}

The matrix $\mathcal{S}_{8}$ is related to matrix $e^{\mathcal{P}_{8}}$ known as
$A056857$ in the Online Encyclopedia of Integer Sequences \citep{OEIS} by the
identity $ e^{\mathcal{P}_{8}}=e\cdot\left(\mathcal{S}_{8}\cdot
\mathcal{P}_{8}\cdot \mathcal{S}_{8}^{-1}\right)$, more connections
involving these matrices can be found in \citep{CHEON200149}; additionally, we report
sine and cosine function applications 
$sin{\mathcal{P}_{8}}$,\,$cos{\mathcal{P}_{8}}$,\,$sin{\mathcal{C}_{8}}$,\,$cos{\mathcal{C}_{8}}$
,\,$sin{\mathcal{S}_{8}}$ and $cos{\mathcal{S}_{8}}$
in the following left-rotated tables, respectively.

%\label{subsec:sines-cosines}
\vspace*{-1cm}
\input{Riordan-matrices-functions/sines-and-cosines.tex}


\section{Jordan canonical form}

\input{Riordan-matrices-functions/jordan-form.tex}
\input{Riordan-matrices-functions/case-study-fibonacci.tex}

\section{Conclusions}

\input{Riordan-matrices-functions/conclusions.tex}






